Bayesian inference of prehistoric demography from multiple proxies: a case study from the Swiss plateau

### 2022-05-23 manuscript/paper.Rmd
Copyedit materials section

The Swiss Plateau (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland's territory and comprises the partly flat but largely hilly area between the Jura Mountains and the Alps. Its average altitude above sea level is between 400 and 600 m. Due to its situation as a basin, it represents a favourable area and is by far the most densely populated region of the Switzerland today. Although the Swiss Plateau is a basin, depending on the region it has a very diverse natural landscape. The Swiss Plateau was shaped by glaciers during the Ice Age, which led to the many lakes that characterise the landscape today, and which provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, as well as a rich pool of sources for vegetation reconstructions by means of pollen analyses. A very active and efficient archaeological research and heritage management also offers a rich repository of archaeological data, known sites and ^14^C data.

@@

In principle, a large number of different data sources can be integrated into the overall model as observations, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created. A non-exhaustive list can be found in the following table \@ref(tab:tableproxies). For our analysis of the Swiss Plateau we used an open land indicator, an aoristic sum of sites based on typological dating, a sum calibration and the number of dendro dated lakeshore settlements in the Three Lakes area. The individual proxies will be discussed in more detail below.

@@

With a mild climate compared to the Jura and the Alps and a dense network of water bodies, the Swiss Plateau is a favourable area for human settlement in Switzerland. From the Neolithic onwards, settlement areas were concentrated along its rivers and lakes [@christianlüthi2009]. Thus, our working region offers on the one hand excellent data for demographic estimation, but on the other hand poses very specific problems for such an undertaking. If we have high-resolution information on the temporal sequence of individual settlements at the lakeside settlements by means of dendro data, this also might cause a research problem with regard to the ^14^C data often used as a proxy.

@@

The dataset we use for the number of dendrodated wetland settlements in the Three Lakes region was collected by Julian Laabs for his PhD thesis [@laabs2019]. Details on the creation of this data series will be published at the referenced location. The time series used here runs from `r abs(min(dendro$date))` to `r abs(max(dendro$date))` BCE, and contains the number of chronologically registered fell phases at individual settlements. This results in a time series that reflects the settlement of the lakeshores in the Neolithic and Bronze Age periods.

@@

The dataset for the ^14^C sum calibration primarily consists of data from the XRONOS database, supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of [@martínez-grau2021]. It contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites. The dates were selected so that their distribution area coincides with the catchment area of the pollen proxy (see also Figure \@ref(fig:c14map)). The dates in the dataset range in ^14^C years from `r max(c14$bp)` to `r min(c14$bp)` BP uncal. This time window extends beyond the study horizon in order to minimise boundary effects, which, however, cannot be completely avoided due to the Hallstatt plateau. But it was precisely to reduce its influence that the time window of the investigation was chosen to be 7000 -- 1000 BCE. The upper limit is set by the post-glacial changes in the pollen spectra, which are hardly associated with human influence before 7000 in the working area, and which would consequently distort the openness indicator. The lower limit is defined by the Hallstatt Plateau (with buffer), which would have direct and indirect research influences on the ^14^C proxy.

We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites. For the creation of the cumulative calibration, the corresponding functions of the R package rcarbon [@crema2021] were used in the standard settings.

@@

In order to add another indicator of archaeological evidence of occupation, we have included the data of the Cantonal Archaeology (Figure \@ref(fig:aoristmap)), and thus the Heritage Management, which are primarily derived from scattered surface finds, and which often have a low dating accuracy. This information is incorporated into our model as a typologically dated aorist time series. The dating accuracy is only in the range of archaeological periods, but the advantage is that we are not bound to the conditions and problems of radiocarbon dating and thus methodological issues of sum calibration can be avoided. Much more, these data provide an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by similar transmission filters and archaeological conditions as the evaluation of ^14^C data. Data from `r nrow(aorist)` sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy in archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.

@@

The natural conditions provided by the many lakes enable not only highly precise dating of archaeological sites, but also a very dense network of pollen analysis. We make use of this fact by generating a supra-regional openness indicator for the vegetation from the pollen data (Figure \@ref(fig:pollensites)). This indicator has the specific advantage that it is not dependent on preservation conditions, as archaeological indicators are. This makes it particularly valuable for indicating or compensating for systematic distortions that result from temporally specific settlement patterns and archaeological preservation conditions.

@@

The utilisation of this proxy is based on the assumption that the higher the population density in an area, the greater the human influence on the natural environment [@lechterbeck2014], and that the agricultural space of an area is closely related to the population density [@zimmermann2004]. Evidence of land clearing in pollen diagrams can therefore provide further indications of population dynamics where humans are the main driver of this process, which is the case in much of Europe. The exact methodological procedure for obtaining this proxy from several, in our case 5, different pollen diagrams of sites mainly in the hinterland of the large Alpine lakes can be found in a previous publication [@heitz2021]. The technical procedure is also documented in the accompanying R-script. The percentage pollen data based on a pollen sum of all terrestrial taxa of the individual sites were combined into one data set by means of a principal component analysis (Figure \@ref(fig:pollenproxy)). Only terrestrial pollen taxa with a frequency of more than 1/3 and, if present, with an average frequency of at least more than 0.1% were selected to reduce potential disturbance by rare species. Cereal pollen was explicitly retained as an important anthropogenic indicator. As each sample is absolutely dated, the data on the x-axis can be plotted against the openness value on the y-axis to obtain a time series time series for land clearing.

@@

Unfortunately, in the Swiss Plateau there is no data from burials that could be usefully incorporated into such this model due to the absence of regular burials for larger parts of prehistory. Formation conditions, in this case most likely C-transformation [@schiffer1987], prevent their use in this study. Nevertheless, we see a very high potential for other regions in the integration of demographic indicators from burial data in order to enlarge the canon of methods and the range of proxies.

The values of the sum calibration, the openness index and the dendrodated settlements were smoothed by means of a moving average with a window of 50 years. Since the aoristic sum already had a very coarse temporal resolution, this was not applied for this measure. The range of the smoothing window corresponded to the sample interval, with which a unified resolution of 50 years was obtained for all proxies as time slices for the model. In addition, all data were restricted to the window of observation of 7000--1000 BCE.

In our following modelling (see also \@ref(observational-model)) we now consider all these proxies as determinants for the change in the number of settlements. In the model, we therefore change the causality and shift the measurement error, which is certainly inherent in each of these indicators, to the process model, in which a Poisson process describes the number of simultaneous settlements. In doing so, we establish a likelihood that indicates how credible the data are, given the model.

@@

>>>>

The Swiss Plateau (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland's territory and comprises the partly flat, but largely hilly area, between the Jura Mountains and the Alps. It averages an altitude of between 400 and 600 m a.s.l. It is a favourable area for settlement and agriculture and is by far the most densely populated region of the Switzerland today. Although the Swiss Plateau is a basin, depending on the region it has a very diverse natural landscape: shaped by glaciers during the Ice Age, the many lakes that characterise the landscape today also provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, as well as a rich pool of sources for vegetation reconstructions by means of pollen analyses. A very active and efficient archaeological research and heritage management also offers a rich repository of archaeological data, known sites and ^14^C data.

@@

Our case study targets the period between 7000–1000 BCE. The lower limit of this window was chosen to avoid the so-called 'Halstatt plateau' on the Northern Hemisphere radiocarbon calibration curve, which would cause difficulties for the ^14^C proxy data. The upper limit was chosen to coincide with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence before 7000 in the working area.

In principle, a large number of different data sources can be integrated into the overall model as proxies, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created. \@ref(tab:tableproxies) provides a non-exhaustive list. For our case study of the Swiss Plateau, we used: a landscape openness indicator; an aoristic sum of sites based on typological dating; a sum calibration; and frequency data for dendro-dated lakeshore settlements in the Three Lakes area.

@@

With a mild climate compared to the Jura and the Alps, and a dense network of water bodies, the Swiss Plateau has always been a favourable area for human settlement in Switzerland. From the Neolithic onwards, settlement areas were concentrated along its rivers and lakes [@christianlüthi2009]. Thus, our working region offers on the one hand excellent data for demographic estimation, but on the other hand poses very specific problems for such an undertaking. If we have high-resolution information on the temporal sequence of individual settlements at the lakeside settlements by means of dendro data, this also might cause a research problem with regard to the ^14^C data often used as a proxy.

@@

The dataset we use for the number of dendro-dated wetland settlements in the Three Lakes region was collected by Julian Laabs for his PhD thesis [@laabs2019]. Details on the creation of this data series will be published at the referenced location. The time series used here runs from `r abs(min(dendro$date))` to `r abs(max(dendro$date))` BCE, and contains the number of chronologically registered fell phases at individual settlements. This results in a time series that reflects the settlement of the lakeshores in the Neolithic and Bronze Age periods.

@@

The dataset for the ^14^C sum calibration primarily consists of data from the XRONOS database (https://xronos.ch), supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of [@martínez-grau2021]. It contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites. The dates were selected so that their distribution area coincides with the catchment area of the pollen proxy (see also Figure \@ref(fig:c14map)). The dates in the dataset range in ^14^C years from `r max(c14$bp)` to `r min(c14$bp)` uncal BP. This time window extends beyond the study horizon in order to minimise boundary effects.

We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites. For the creation of the cumulative calibration, the corresponding functions of the R package rcarbon [@crema2021] were used with their default settings.

@@

To add another archaeological indicator of occupation, we include relative dating information obtained from the Swiss cantonal archaeology/heritage management authorities (Figure \@ref(fig:aoristmap)). These are primarily derived from scattered surface finds, which often have a low dating accuracy. We incorporate this data into our model as a typologically dated aorist time series. The dating accuracy is only in the range of archaeological periods, but the advantage is that we are not bound to the conditions and problems of radiocarbon dating and thus methodological issues of sum calibration can be avoided. Furthermore, this data provides an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by similar transmission filters and archaeological conditions as the evaluation of ^14^C data. Data from `r nrow(aorist)` sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy offered by archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.

@@

The natural conditions of the lakes of the Swiss plateau enable not only highly precise dating of archaeological sites, but also a very dense network of pollen analysis. We make use of this fact by generating a supra-regional openness indicator for the vegetation from the pollen data (Figure \@ref(fig:pollensites)). This proxy has the specific advantage that it is not dependent on preservation conditions, as archaeological indicators are. This makes it particularly valuable for indicating or compensating for systematic distortions that result from temporally specific settlement patterns and archaeological preservation conditions.

@@

The utilisation of this proxy is based on the assumption that the higher the population density in an area, the greater the human influence on the natural environment [@lechterbeck2014], and that the extensiveness of agricultural activity in an area is related to human population density [@zimmermann2004]. Evidence of land clearance in pollen diagrams can therefore provide further indications of population dynamics where humans can be assumed to be the main driver of this process, which is the case in much of Europe. The full procedure for deriving this proxy from several different pollen diagrams is detailed in a previous publication [@heitz2020]. In this study, five pollen diagrams from sites mainly in the hinterland of the large Alpine lakes were used. The technical steps are also documented in the accompanying R compendium. The percentage pollen data, based on a pollen sum of all terrestrial taxa of the individual sites, was combined into one data set by means of a principal component analysis (Figure \@ref(fig:pollenproxy)). Only terrestrial pollen taxa with a frequency of more than one third and, if present, with an average frequency of at least 0.1% were selected to reduce potential disturbance by rare species. Cereal pollen was explicitly retained as an important anthropogenic indicator. As each sample is absolutely dated, the data on the x-axis can be plotted against the openness value on the y-axis to obtain a time series time series for land clearance.

@@

## Other proxies

Unfortunately, there is no burial data available the Swiss plateau that could be usefully incorporated into such this model due to the absence of regular burials for larger parts of prehistory. Formation conditions, in this case most likely C-transformation [@schiffer1987], prevent their use in this study. Nevertheless, we see a very high potential for other regions in the integration of demographic indicators from burial data in order to enlarge the canon of methods and the range of proxies.

@@

The sum calibration, openness and the dendro-dated settlement data was smoothed by means of a moving average with a window of 50 years. Since the aoristic sum already had a very coarse temporal resolution, this was not applied for this measure. The range of the smoothing window corresponded to the sample interval, with which a unified resolution of 50 years was obtained for all proxies as time slices for the model. In addition, all data were restricted to the window of observation of 7000--1000 BCE.

In the construction of our 'observational model', we considered all these proxies as informative of the number of settlements in the Swiss plateau. Consequently, we shift the causality and consideration of measurement error, which is certainly inherent in each of these proxies, to a preceding 'process model', in which a Poisson process describes the number of simultaneous settlements. In doing so, we establish a likelihood that estimates how credible the data is, given the model.
